---
title: "Exercise 5"
author: "Philipp Drebes"
date: "`r format(Sys.Date(), '%d.%m.%Y')`"
header-includes:
   - \usepackage{siunitx}
output:
  pdf_document:
    keep_tex: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Task 1: Hierarchical Clustering

The data `empl2.dat` contains the rate of employment (in percent, rounded) in 9 different sectors in 10 European countries in the year 1979. The countries are the following ones: B Belgium, CH Switzerland, CS Czechoslovakia, D BRD, GB Great Britain, GR Greece, H Hungary, S Sweden, TR Turkey, YU Yugoslavia.

a)  First look at the data using a scatterplot. Can you find clusters by eye?

```{r}
empl <- read.table('../data/empl2.dat', header=T)
labempl <- rownames(empl)
pairs(empl, panel=function(x,y) text(x,y, labels=labempl,xpd=T))
```

b)  Calculate the Euclidean distances between the states. Which two states are first combined into a cluster?

```{r}
empl.dist <- dist(empl)
round(empl.dist, 2)
min(empl.dist)
```

c)  Carry out a hierarchical cluster analysis by hand using the "single linkage"-method. Hints: Recall how the distance between two clusters is defined in the single linkage method and based on this, think of how the algorithm proceeds.

```{r}

```

d)  Carry out the previous cluster analysis using the function `hclust()`. Verify your result of c) by comparing the first five steps.

```{r}
empl.single <- hclust(empl.dist, method='single')
plot(empl.single)
```

e)  Carry out the cluster analysis with the same distances but with the methods average and complete. Compare the dendrograms of all three methods.

```{r}
empl.avg <- hclust(empl.dist, method='average')
plot(empl.avg)

empl.comp <- hclust(empl.dist, method='complete')
plot(empl.comp)
```

f)  Group the states into k clusters. Choose for instance k = 4 or k = 3. Compare the different methods from d) and e). Also plot an MDS-plot and mark the observed groups of states with colors (for one k and one method).

```{r}
# Classification in three groups using Single Linkage 
r.3scl <- cutree(sing.empl, k=3)
split(labempl, r.3scl)
r.mds <- cmdscale(...)
plot(r.mds, type = "n", asp=1, main = "Single clustering, MDS coordinates")
text(r.mds, labempl, col = 1 + r.3scl)
```
