---
title: "Exercise 4"
author: "Philipp Drebes"
date: "`r format(Sys.Date(), '%d.%m.%Y')`"
header-includes:
   - \usepackage{siunitx}
output:
  pdf_document:
    keep_tex: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Task 1: MDS for road distances

The data `ch-dist.dat` consists of distances between 12 Swiss cities. We would like to determine the location of these cities. The distances are given as fastest road routes, not shortest routes nor air distances.

a)  Determine the "coordinates" of these cities using classical MDS and plot the resulting points. Are the locations of the cities approximately equal to their locations on the map of Switzerland? For comparison you may have to reflect or transform the coordinates.

```{r}
cities <- read.table("../data/ch-dist.dat", header=TRUE)
head(cities)

cities.mds <- cmdscale(cities, k = 2, eig = TRUE)
cities.coords <- cities.mds$points
plot(-cities.coords[,1], -cities.coords[,2], type = 'n')
text(-cities.coords[,1], -cities.coords[,2], labels = row.names(cities.coords))
```

b)  The goal of this exercise is to investigate how well the MDS reconstruction approximates the true distances by comparing the original distances with the distances from the MDS solution. Plot the points using the distances of the MDS solution as y-axis and the original distances as the x-axis. Which range of points of the distances are well approximated?

```{r}
mds.dist <- as.vector(as.matrix(dist(cities.coords, diag=T, upper=T)))
orig.dist <- as.vector(as.matrix(cities))

fit <- lm(mds.dist ~ orig.dist)
plot(orig.dist, mds.dist)
abline(fit, col = 4)
```

The points below 100km and 300km distance, are well approximated.

## Task 2: MDS for country statistics

We collected data on 180 countries from the CIA World Fact Book1 on the variables: area, population size, life expectancy, fertility rate, GDP (purchasing power parity), GDP growth, unemployment rate. The values are either 2011 or 2012 estimates.

a)  Load the data `countries.rda` and look at the summary of the data set. Scale the data such that all variables have mean 0 and variance 1.

```{r}
load(file="../data/countries.rda")
summary(countries)

centered.countries <- scale(countries, center = TRUE, scale = TRUE)
summary(centered.countries)
```

b)  Compute Euclidean distances and perform multidimensional scaling (MDS). What is the number of dimensions that should be used?

```{r}
countries.dist <- dist(centered.countries)

countries.mds <- cmdscale(countries.dist, k = 2, eig = T)
countries.coords <- countries.mds$points



num_data <- attr(countries.dist, "Size")

## Choosing the dimension
mds <- cmdscale(countries.dist, k = num_data - 1, eig = T)
ev <- mds$eig

## Choose number of dimensions
cumsum(ev) / sum(ev)
cumsum(abs(ev)) / sum(abs(ev))
cumsum(ev^2) / sum(ev^2)

## Scree plot
barplot(Eigenvalue ~ Coordinate,
        data = data.frame(Eigenvalue=ev[1:(num_data-1)],Coordinate=1:(num_data-1)))
```

c)  Refit with two dimensions. Plot dots for all the countries and highlight Switzerland.

```{r}
countries.mds <- cmdscale(dist(centered.countries), k = 2, eig = T)
countries.coords <- countries.mds$points
```

d)  Find the three countries that are most similar to Switzerland.

```{r}
plot(countries.coords)
identify(countries.coords[,1], countries.coords[,2], labels = rownames(countries.coords))
```

e)  Perform PCA and plot the first two principal components.

```{r}

```

f)  Why is it necessary to scale the data if we want to apply multidimensional scaling (MDS)?

```{r}

```

## Task 3

The file `mnist_2k.Rdata` contains images and labels of 2000 randomly chosen hand-written digits, from the MNIST data set. Load the data (read in with `load()`) and familiarize yourself with the data.

a)  Visualize one image.

```{r}
## Load data
load(file="../data/mnist_2k.Rdata")

table(x$label) ## Frequencies of of digigts
names(x)[1:10]

## Create colors for digits (used below in plots)
coldigits=rainbow(10)[x[,1]+1]

## Visualize one image
N <- 28 # width/height of mnist images in pixels
i <- 620 # pick number of image to visualize
image <- matrix(as.numeric(x[i, -1]), ncol = 28, byrow = TRUE)
heatmap(image, Rowv = NA, Colv = NA, revC = TRUE, asp = 1, scale = "none")
```

b)  Run PCA for the data. Which percentage of the total variance is explained by the first two principal components? Create a 2D plot with the scores and use colors to mark the points according to the corresponding labels (i.e., the digits).

```{r}
digits.pca <- princomp(x[,-1])
plot(digits.pca)
cumsum(digits.pca$sdev[1:20]^2 / sum(digits.pca$sdev^2))

plot(digits.pca$score[,1], digits.pca$score[,2],col=coldigits,xlab ="",ylab="")
legend("bottomright",legend=0:9,col=rainbow(10),bty = "n",pch=1,pt.lwd=2)
```

c)  Run t-SNE to obtain a two-dimensional representation of the data. Compare to the results from PCA and comment on the differences.

```{r}
library(Rtsne)

set.seed(9898)

digits.tsne <- Rtsne(x[, -1], dims = 2, perplexity = 20, max_iter = 1000)
plot(digits.tsne$Y[,1], digits.tsne$Y[,2], xlab = "", ylab = "", col = coldigits)
legend("topleft", legend = 0:9, col = rainbow(10), bty = "n", pch = 1, pt.lwd = 2)
```

## Task 4

$$
\frac{1}{8} \cdot (1 + 0 + 0 + 1 + 1 + 1 + 1.5/4.1 + 1.8/4.3) = 0.598
$$
